---
title: "shinyAssistantUI overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{shinyAssistantUI overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What shinyAssistantUI provides

`shinyAssistantUI` brings the React Web surface from
[assistant-ui](https://www.assistant-ui.com/) into a Shiny htmlwidget. It keeps the
assistant-ui interaction model while using Shiny's normal client/server channel and R
callbacks for application logic.

The package is backend-agnostic. You can supply a small R handler, use one of the packaged
integrations, or build a handler around another service.

| assistant-ui concept | shinyAssistantUI implementation |
|---|---|
| Surface | React Web inside an htmlwidget |
| Runtime | A custom `ExternalStoreRuntime` owned by the widget |
| Transport | Shiny inputs and custom messages over its WebSocket |
| Server/runtime | A backend-agnostic R handler |
| Included integrations | ClaudeAgentSDK, ellmer, and custom/codeagent handlers |
| Native or terminal surfaces | Outside this package's Shiny Web scope |
| AG-UI, A2A, or other runtime adapters | Not required or installed |

This is the project's mapping of the first assistant-ui Getting Started page,
**Documentation**. It is a positioning guide rather than a separate JavaScript runtime.

## Minimal Shiny app

A handler receives the user's text and callbacks used to stream or complete the response.
The example below has no external AI dependency.

```{r minimal-app, eval=FALSE}
library(shiny)
library(shinyAssistantUI)

ui <- assistantUIPage(
  assistantUIOutput("chat", height = "100%"),
  title = "AI Assistant"
)

server <- function(input, output, session) {
  assistantUIServer(
    "chat",
    handler = function(message, on_chunk, on_done, on_error) {
      on_chunk("Hello! You said: ")
      on_chunk(message)
      on_done()
    }
  )
}

shinyApp(ui, server)
```

## Choose the R-side integration

Use the smallest integration that matches the application:

- Pass a function directly to `assistantUIServer()` for a custom backend.
- Use `make_claude_handler()` when ClaudeAgentSDK session and control features are needed.
- Use `make_ellmer_handler()` for an ellmer chat backend.
- Use the codeagent helpers when that execution model is explicitly required.

The UI is not coupled to any one of these choices. Each integration translates backend events
into the same widget-facing stream, tool, approval, session, and status messages.

## Runtime flow

```text
React composer
  -> Shiny input
  -> assistantUIServer()
  -> R handler or packaged integration
  -> Shiny custom messages
  -> ExternalStoreRuntime
  -> assistant-ui React components
```

assistant-ui owns the React interaction primitives. `shinyAssistantUI` owns the bridge that
turns Shiny events into runtime state and turns R callbacks into incremental UI updates.

## Scope decisions

assistant-ui also documents React Native, terminal surfaces, hosted transports, and several
runtime adapters. They are valid choices for other applications, but they are not prerequisites
for this package's React Web + Shiny architecture. In particular, this documentation foundation
does not install A2UI or AG-UI packages and does not replace the existing custom runtime.

For the ordered review record and deferred pages, see
[Upstream documentation alignment](upstream-alignment.html).
